An Algebraic Representation of Wireless Workload

### 2008-02-08 paper.tex
Done with results; working on discussion.

@@

The story of network analysis and modeling is one of variety. Assumptions of ``sameness'' have fallen repeatedly. ...
The matrix model gives new insight into

What are the ramifications of these results? The discovery that the most commonly used traffic model for wireless networks drastically misrepresents important performance metrics may shed some light on the lack of trust in results from wireless simulations. It is now well established that network usage behavior---both mobility, and, with this research, traffic patterns---have an impact on network performance that cannot be ignored. Even experimental deployments cannot avoid the need for more realistic traffic workload models. While using a real, physical network successfully sidesteps simulation problems below the application layer, without realistic traffic models, reliable, meaningful performance predictions remain beyond our reach.

\subsection{Generality of Results}

% TODO: talk about other synthetic models not considered here.
% The point of our analysis is not to show that no synthetic models can be accurate, but rather, that the commonly used models, and obvious variations thereof, are not accurate.

The most significant limitation on the generality of this analysis is that it is based entirely on a single data set from \caps{IETF}60---albeit a large and varied one. It is possible that traffic in this trace happens to produce network performance that is unusually dissimilar to standard traffic models. This data set, however, represents a highly heterogeneous collection of network usage behaviors, from slow and steady off-peak usage, to extremely heavy peak usage: over 800 users, 33 thousand flows, and 1 million packets in a single 10-minute trace segment. Despite the broad variety of behaviors, the results are consistent: in all types of usage scenarios, simplistic traffic models, like uniform \caps{CBR}, systematically skew important performance measurements at all levels of the network. While the precise results for other data sets might differ, it is very unlikely that \caps{CBR} traffic models will happen to accurately reproduce realistic performance in other experiments. This paper provides strong evidence that better traffic models are needed for performance evaluations.

@@

What would better traffic models look like? How can we create them? One possible approach is to use actual traffic traces as we have done. This approach is unsatisfactory, however, because it provides the experimenter with almost no control over experiments. Synthetic models have parameters, which can be tweaked as necessary---adjusting, for example, the number of active nodes in a simulation, without affecting other parameters. Traffic traces, on the other hand, must be used without significant alteration if they are to actually provide the desired realism. %The ``messiness'' of the performance comparison from trace data in Figure~\ref{fig:mac-control-overhead-cmp} illustrates why using traces directly is not ideal: each data point differs not only in the number of nodes shown on the $x$-axis, but also in other dimensions, such as the number of flows and packets, and the average flow duration. The result is a highly noisy comparison, affected by many unseen parameters. Only by applying a local smoothing algorithm are trends somewhat elucidated.

Instead of using trace data directly, it should be possible to configure a synthetic traffic model based on observations from a real data set, and then run side-by-side simulations using the synthetic model and the real data, producing statistically equivalent performance results. This is precisely what our definition of sufficient realism entails. The work in this paper provides the tools to measure how close to this ideal a model is and in what areas it needs improvement. Without this feedback, any improvements in realism are purely guesswork. Our breakdown of traffic behavior into three orthogonal levels also allows the problem to be approached in smaller pieces, rather than being solved all at once.

>>>>

Finally, we turn to the regularized models. These demonstrate neither exceptionally good nor exceptionally bad behavior. That the regularized models should be similar to the \model{full marginal} model but somewhat more realistic is entirely expected: regularization, the reader will recall, is a restricted form of marginal modeling, where each group of flows has its own marginal distribution. In these cases, the flows are grouped by source node or by destination node. The mixed case uses the average behavior of the other two cases. It is interesting to note that the mixed case is universally worse than either the source or destination case. Clearly, hybridizing the regularization here is not an effective technique. We believe that the cause of this effect is like mixing too many colors yielding an unpleasant shade of brown: any of the colors individually would be better than the non-descript average. As a more general conclusion, arbitrarily mixing model may not always be a good idea, despite the ease with which it is done in the matrix representation of workload behavior.

@@

%The story of network analysis and modeling is one of variety. Assumptions of ``sameness'' have fallen repeatedly. ...
%The matrix model gives new insight into

What are the ramifications of these results? In our previous work, we had already demonstrated the drastic distortions that are introduced by uniformity assumptions made by the most common traffic models. This work serves to further critique established common assumptions in traffic modeling. The strongest message to be taken away is that the aspects of traffic behavior of both nodes and flows \textit{must} be considered jointly. It is not sufficient to continue modeling marginal distributions of network properties. These marginal models do not lend them selves easily to traffic generation in the first place, and when marginality assumptions are applied to trace traffic, the result is a severe distortion of important network performance characteristics. It is now well established that network traffic patterns have an impact on network performance that cannot be ignored. Even experimental deployments cannot avoid the need for more realistic traffic workload models. While using a real, physical network successfully sidesteps simulation problems below the application layer, without realistic traffic models, reliable, meaningful performance predictions remain beyond our reach.

%\subsection{Generality of Results}

%% TODO: talk about other synthetic models not considered here.
%% The point of our analysis is not to show that no synthetic models can be accurate, but rather, that the commonly used models, and obvious variations thereof, are not accurate.

%The most significant limitation on the generality of this analysis is that it is based entirely on a single data set from \caps{IETF}60---albeit a large and varied one. It is possible that traffic in this trace happens to produce network performance that is unusually dissimilar to standard traffic models. This data set, however, represents a highly heterogeneous collection of network usage behaviors, from slow and steady off-peak usage, to extremely heavy peak usage: over 800 users, 33 thousand flows, and 1 million packets in a single 10-minute trace segment. Despite the broad variety of behaviors, the results are consistent: in all types of usage scenarios, simplistic traffic models, like uniform \caps{CBR}, systematically skew important performance measurements at all levels of the network. While the precise results for other data sets might differ, it is very unlikely that \caps{CBR} traffic models will happen to accurately reproduce realistic performance in other experiments. This paper provides strong evidence that better traffic models are needed for performance evaluations.

@@

What would better traffic models look like? How can we create them? One possible approach is to use actual traffic traces as we have done. This approach is unsatisfactory, however, because it provides the experimenter with almost no control over experiments. Synthetic models have parameters, which can be tweaked as necessary---adjusting, for example, the number of active nodes in a simulation, without affecting other parameters. Traffic traces, on the other hand, must be used without significant alteration if they are to actually provide the desired realism.

On the other hand, we have discredited both uniform and marginal modeling as viable approaches that can fully address the workload modeling requirements of the wireless research community. This, we believe, is where the General Matrix Model plays a vital role in the development of better models.

%The ``messiness'' of the performance comparison from trace data in Figure~\ref{fig:mac-control-overhead-cmp} illustrates why using traces directly is not ideal: each data point differs not only in the number of nodes shown on the $x$-axis, but also in other dimensions, such as the number of flows and packets, and the average flow duration. The result is a highly noisy comparison, affected by many unseen parameters. Only by applying a local smoothing algorithm are trends somewhat elucidated.

%Instead of using trace data directly, it should be possible to configure a synthetic traffic model based on observations from a real data set, and then run side-by-side simulations using the synthetic model and the real data, producing statistically equivalent performance results. This is precisely what our definition of sufficient realism entails. The work in this paper provides the tools to measure how close to this ideal a model is and in what areas it needs improvement. Without this feedback, any improvements in realism are purely guesswork. Our breakdown of traffic behavior into three orthogonal levels also allows the problem to be approached in smaller pieces, rather than being solved all at once.
